Topic modeling on EU and the USA research projects

### 2017-06-25 thesis_report/draft_thesis.tex
Comparison of the two datasets with JSD

where $k$ is the number of the topic id, $y$ is the year between 1994-2016. The
topic proportion of each topic per year equals to the sum of the number of the
documents assigned to this topic during the year, normalized by the number of
the documents-projects  occurring in this year.

In figure X we plot the topic distribution over time. We can easily observe that
the topic 6, which refers to STEM education, is becoming a trend in the United
States over the years, while most of the other topics' evolution doesn't show up
such a high increase. Topics 5 and 8, which refer to Chemistry \& Materials
science and Molecular biology, respectively, show a small decrease over the
years. Topics 9 and 0 (Mechanical Engineering, Material Science and High energy
physics) show a small increase over the last 5 years. Mathematics' evolution
over time (Topic 1) looks stable on average and shows up as a popular subject
over the last 22 years, while Social Sciences and International Research (Topic
3 and 4) show a small decrease over the last 3 years.  Molecular biology and
Genetics (Topic 2) has an increasing popularity from 1994 to 2001 and later is
starting to stabilize. Climatology, Atmospheric and Ocean sciences look stable
the first years, while from 2006 up to date it is gradually becoming less
popular over the years.

In general, the STEM (science, technology, engineering and math) related
research and education is becoming more and more popular, which is not
surprising, since the National Science Foundation itself estimates that 80/% of
the jobs available during the next decade will require math and science skills.
Even the concept of STEM was first introduced by Judith A. Ramaley, the former
director of the National Science Foundation's education and human resources
division.
\begin{center}
\begin{figure}
\includegraphics[width=1.0\textwidth]
{figs/topic-evolution-usa.png}
\caption{Topics over time in USA}
\end{figure}
\end{center}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]
{figs/heatmaps/heatmapFP4.png}
\caption{Heatmap of the similarity of the two regions during 1994-1997}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]
{figs/heatmaps/heatmapFP5.png}
\caption{Heatmap of the similarity of the two regions during 1998-2001}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]
{figs/heatmaps/heatmapFP6.png}
\caption{Heatmap of the similarity of the two regions during 2002-2006}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]
{figs/heatmaps/heatmapFP7.png}
\caption{Heatmap of the similarity of the two regions during 2007-2013}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]
{figs/heatmaps/heatmapH2020.png}
\caption{Heatmap of the similarity of the two regions during 2014-2017}
\end{figure}

@@

\begin{center}
\begin{figure}[h]
\begin{tabular}{rcl}
\subfloat[Quality Management Systems]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic0_7000.png}} &
\subfloat[Agriculture, Fisheries and Food]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic1_7000.png}}&
\subfloat[Molecular biology, Genetics]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic2_7000.png}}\\
\subfloat[EU transport]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic3_7000.png}}&
\subfloat[Digital systems]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic4_7000.png}} &
\subfloat[Materials processing]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic5_7000.png}}\\
\subfloat[Genetics]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic6_7000.png}} &
\subfloat[Policy for environmental sustainability]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic7_7000.png}}&
\subfloat[Energy technologies]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic8_7000.png}}\\
\subfloat[Applications in mobile networks]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic9_7000.png}}&
\end{tabular}
\caption{Wordclouds of topics of the EU FP4 (1994-1998)}
\end{figure}
\end{center}

\begin{center}
\begin{figure}[h]
\begin{tabular}{rcl}
\subfloat[Management software]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic0_7000.png}} &
\subfloat[Hydropower / Energy production]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic1_7000.png}}&
\subfloat[Information technologies]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic2_7000.png}}\\
\subfloat[Quality control systems]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic3_7000.png}}&
\subfloat[Health care / Medical genetics]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic4_7000.png}} &
\subfloat[Environmental sustainability and policies]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic5_7000.png}}\\
\subfloat[Molecular biology (genes)]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic6_7000.png}} &
\subfloat[Optical quantum computing]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic7_7000.png}}&
\subfloat[Education in Nuclear physics]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic8_7000.png}}\\
\subfloat[Education, training]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic9_7000.png}}&
\end{tabular}
\caption{Wordclouds of topics of the EU FP5 (1998-2002)}
\end{figure}
\end{center}

>>>>

where $k$ is the number of the topic id, $y$ is the year between 1994-2016. The
topic proportion of each topic per year equals to the sum of the number of the
documents assigned to this topic during the year, normalized by the number of
the documents-projects  occurring in this year.
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]
{figs/topic-evolution-usa.png}
\caption{Topics over time in USA}
\end{figure}
\end{center}
In figure 6 we plot the topic distribution over time. We can easily observe that
the topic 6, which refers to STEM education, is becoming a trend in the United
States over the years, while most of the other topics' evolution doesn't show up
such a high increase. Topics 5 and 8, which refer to Chemistry \& Materials
science and Molecular biology, respectively, show a small decrease over the
years. Topics 9 and 0 (Mechanical Engineering, Material Science and High energy
physics) show a small increase over the last 5 years. Mathematics' evolution
over time (Topic 1) looks stable on average and shows up as a popular subject
over the last 22 years, while Social Sciences and International Research (Topic
3 and 4) show a small decrease over the last 3 years.  Molecular biology and
Genetics (Topic 2) has an increasing popularity from 1994 to 2001 and later is
starting to stabilize. Climatology, Atmospheric and Ocean sciences look stable
the first years, while from 2006 up to date it is gradually becoming less
popular over the years.

In general, the STEM (science, technology, engineering and math) related
research and education is becoming more and more popular, which is not
surprising, since the National Science Foundation itself estimates that 80/% of
the jobs available during the next decade will require math and science skills.
Even the concept of STEM was first introduced by Judith A. Ramaley, the former
director of the National Science Foundation's education and human resources
division.

@@

In LDA, a topic is represented as a probability distribution over words of the
vocabulary. The words with the highest probability are the ones that most likely
describe the topic, based on the model inference. In order to compare these
topics quantitatively, one can apply measures that compare probability
distributions, like KL-divergence or Jensen Shannon divergence (JSD). To make
this computation more concrete, we will give a basic definition:

Two topics (topic 1 and topic 2) are represented as normalized counts, or
multinomial probability distributions.
\begin{itemize}
\item $P=(p_1, p_2, p_3,…, p_N)$ the probability distribution of the words in topic 1.
\item $Q=(q_1, q_2, …, q_N)$ the multinomial distribution of the words in topic 2.
\item $N = $total number of words in the vocabulary.
\end{itemize}

Since P and Q are probability distributions, it means that 
$\sum_{i=1}^{N} p_i = \sum_{i=1}^{N} q_i = 1$. The The Jensen–Shannon 
divergence (JSD) is a
symmetrized and smoothed version of the Kullback–Leibler divergence
$D(P \mid\mid Q)$. It quantifies how "distinguishable" two distributions are
from each other. It is defined by

\begin{equation}
JSD(P \mid\mid Q) = \frac{1}{2}D(P \mid\mid M) + \frac{1}{2}D(Q \mid\mid M)
\end{equation}
where $M = \frac{1}{2}(P + Q)$ and $D(P \mid\mid Q) = \sum_{i}^{N} p_i \ln\frac{p_i}{q_i}$.

In our case, we will use JSD cause is a symmetric formula, so that 
$sim(Q,P) = sim(P,Q)$. we want to calculate the similarity of the topics of the 
EU and the USA dataset per framework programme, one by one. Since the topics 
come from different datasets, it means that their topic probability distributions 
have different variables-words and different number of variables. To deal with this
issue, we run the model for the two datasets separately and keep the topics as
word probability distributions. Then for each pair of the two set of topics, we
add to each of the topic-word distributions, the words that are missing and
exist in the other distribution, with probability equals to zero. Then we apply
the JSD to each pair of topics (see compare FPs.py in the Appendix, for the
implementation).

The figures 7-11 visualize the heatmaps of the similarity of the two regions
during the FP4-H2020. Rows represent topics from the EU dataset and 
columns topics from the USA dataset. The values of the JSD computation which are 
close to zero (lighter color of the box) suggest more similarity within topics.
According to the heatmaps, the topics that have high similarity between the 
EU and the USA are:
\begin{itemize}
\item FP4 (1994-1997)
\begin{itemize}
\item[*] Software engineering
\item[*] Materials
\end{itemize}
\item FP5 (1998-2001)
\begin{itemize}
\item[*] Energy production
\item[*] Education in STEM
\item[*] Mathematics
\end{itemize}
\item FP6 (2002 - 2006)
\begin{itemize}
\item[*] Information systems
\item[*] Molecular biology / Genetics
\item[*] Chemistry \& Materials
\end{itemize}
\item FP7 (2007-2013)
\begin{itemize}
\item[*] Chemistry \& Materials
\item[*] High energy physics
\item[*] Molecular biology / Genetics
\end{itemize}
\item H2020 (2014-2017)
\begin{itemize}
\item[*] Information systems
\item[*] Materials
\item[*] Physics
\item[*] Economic \& Social policy
\end{itemize}
\end{itemize}
\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]
{figs/heatmaps/heatmapFP4.png}
\caption{Heatmap of the similarity of the two regions during 1994-1997}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]
{figs/heatmaps/heatmapFP5.png}
\caption{Heatmap of the similarity of the two regions during 1998-2001}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]
{figs/heatmaps/heatmapFP6.png}
\caption{Heatmap of the similarity of the two regions during 2002-2006}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]
{figs/heatmaps/heatmapFP7.png}
\caption{Heatmap of the similarity of the two regions during 2007-2013}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]
{figs/heatmaps/heatmapH2020.png}
\caption{Heatmap of the similarity of the two regions during 2014-2017}
\end{figure}
\vfill

@@

\textbf{compare FPs.py}

\lstinputlisting[language=Python]{code/compare_FPs.py}

\begin{center}
\begin{figure}[h]
\begin{tabular}{rcl}
\subfloat[Quality Management Systems]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic0_7000.png}} &
\subfloat[Agriculture, Fisheries and Food]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic1_7000.png}}&
\subfloat[Molecular biology, Genetics]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic2_7000.png}}\\
\subfloat[EU transport]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic3_7000.png}}&
\subfloat[Digital systems]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic4_7000.png}} &
\subfloat[Materials processing]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic5_7000.png}}\\
\subfloat[Genetics]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic6_7000.png}} &
\subfloat[Policy for environmental sustainability]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic7_7000.png}}&
\subfloat[Energy technologies]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic8_7000.png}}\\
\subfloat[Applications in mobile networks]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP4_wordclouds/EUFP4_topic9_7000.png}}&
\end{tabular}
\caption{Wordclouds of topics of the EU FP4 (1994-1998)}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[h]
\begin{tabular}{rcl}
\subfloat[Management software]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic0_7000.png}} &
\subfloat[Hydropower / Energy production]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic1_7000.png}}&
\subfloat[Information technologies]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic2_7000.png}}\\
\subfloat[Quality control systems]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic3_7000.png}}&
\subfloat[Health care / Medical genetics]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic4_7000.png}} &
\subfloat[Environmental sustainability and policies]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic5_7000.png}}\\
\subfloat[Molecular biology (genes)]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic6_7000.png}} &
\subfloat[Optical quantum computing]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic7_7000.png}}&
\subfloat[Education in Nuclear physics]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic8_7000.png}}\\
\subfloat[Education, training]{\includegraphics[width=0.35\textwidth]{figs/eu_topics/eu_FP5_wordclouds/EUFP5_topic9_7000.png}}&
\end{tabular}
\caption{Wordclouds of topics of the EU FP5 (1998-2002)}
\end{figure}
\end{center}
